---
title: "EDA Project"
author: "Sydney"
date: "2026-09-20"
output: html_document
---

#Set up your github connection with your r studio
```{r}
install.packages(c("usethis", "gitcreds"))
library(usethis)
use_git_config(user.name = "YourGitHubUsername", user.email = "your.email@example.com")
usethis::create_github_token()
gitcreds::gitcreds_set()
usethis::use_git()
usethis::use_github()

```
#Libraries
```{r}
#if you need to install a library use this line -- install.packages() 
library(readr)
library(tidyr)
library(dplyr)
library(naniar) #makes visual representation of NA values
library(ggplot2)
#used to disable scientific notation for the entire document(below)
#options(scipen = 999)
```


#getting the data
```{r}
# put this code in your block n modify it to fit your needs like file path & columns

project.dir <- 
  "C:\Users\sydne\Desktop\DS_Fall26\DMungingExpAnalysis\Group Project 1"
# (change to your specific path up to 'Projects')
dataset.dir <- file.path(project.dir, "Datasets",
                         "College_Scorecard_Raw_Data_06102026", 
                         "College_Scorecard_Raw_Data_06032026") #the route to get to all the files
outputs.dir <- file.path(project.dir, "Output")
setwd(project.dir)
list.files(dataset.dir, pattern = "MERGED") #check to see if you're looking at the right files

year9_10 <- read_csv(
  file.path(dataset.dir, "MERGED2009_10_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)
year14_15 <- read_csv(
  file.path(dataset.dir, "MERGED2014_15_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)
year20_21 <- read_csv(
  file.path(dataset.dir, "MERGED2020_21_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)

```


#Reuel's block 
Columns important to my question "how have post-graduation earnings changed over the same period?"
# 1. Identifiers & Matching
  "UNITID",            # Unique institution ID
  "INSTNM",            # School name
  
  # 2. Institutional Type & Filtering 
  "CONTROL",           # 1 = Public, 2 = Private Nonprofit, 3 = Private For-Profit
  "PREDDEG",           # Predominant degree (3 = Predominantly Bachelor's)
  
# 3. Post-Graduation Earnings (Primary Response Variables)
  "MD_EARN_WNE_P10",   # Median earnings of students working & not enrolled, 10 yrs post-entry
  "MD_EARN_WNE_P6",    # Median earnings of students working & not enrolled, 6 yrs post-entry
  "GT_25K_P10",        # Percentage earning > $25k (threshold of a high school graduate)
  
# 4. Context for Financial Investment
  "DEBT_MDN",          # Median cumulative student debt at graduation
  "NPT4_PUB",          # Average net price for public institutions (Title IV students)
  "NPT4_PRIV"
```{r}
project.dir <- "C:/Projects" 
dataset.dir <- file.path(project.dir, "Datasets",
                         "College_Scorecard_Raw_Data_06102026", 
                         "College_Scorecard_Raw_Data_06032026") #the route to get to all the files
outputs.dir <- file.path(project.dir, "Output")
setwd(project.dir)
list.files(dataset.dir, pattern = "MERGED")

year9_10 <- read_csv(
  file.path(dataset.dir, "MERGED2009_10_PP.csv"),
   col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)
year14_15 <- read_csv(
  file.path(dataset.dir, "MERGED2014_15_PP.csv"),
  col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)
year20_21 <- read_csv(
  file.path(dataset.dir, "MERGED2020_21_PP.csv"),
 col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)
#same column name will be necessary for looking at models/graphs
year9_10$year <- "2009-2010"
year14_15$year <- "2014-2015"
year20_21$year <- "2020-2021"
all_years <- rbind(year9_10, year14_15, year20_21)
head(all_years)

```

#assessing data | fixing data types | NA values
```{r}
#visual representation of missing data
gg_miss_var(all_years)

#checking data types
class(all_years$MD_EARN_WNE_P10)
all_years$MD_EARN_WNE_P10 <- as.numeric(all_years$MD_EARN_WNE_P10)

class(all_years$GT_25K_P10)
all_years$GT_25K_P10 <- as.numeric(all_years$GT_25K_P10)

class(all_years$MD_EARN_WNE_P6)
all_years$MD_EARN_WNE_P6 <- as.numeric(all_years$MD_EARN_WNE_P6)

class(all_years$DEBT_MDN)
all_years$DEBT_MDN <- as.numeric(all_years$DEBT_MDN)

class(all_years$NPT4_PUB)
all_years$NPT4_PUB <- as.numeric(all_years$NPT4_PUB)

class(all_years$NPT4_PRIV)
all_years$NPT4_PRIV <- as.numeric(all_years$NPT4_PRIV )

#Fixing NA values

# combine the values because public UNIs can't give NPT$ for Priv UNIs
all_years <- all_years %>%
  mutate(NPT4_combined = coalesce(NPT4_PUB, NPT4_PRIV))

#government stopped recording this variable (GT_25K_P10) past 2012 so that explains the NA values  
```


## Checking correlation with variables ##
```{r}
#checking variables with response variable
cor.test(all_years$CONTROL,all_years$PREDDEG)
#comparing 
cor.test(all_years$PREDDEG,all_years$MD_EARN_WNE_P6) #.40 
cor.test(all_years$PREDDEG,all_years$GT_25K_P10) #.55
cor.test(all_years$PREDDEG,all_years$MD_EARN_WNE_P10) #.46

cor.test(all_years$PREDDEG,all_years$DEBT_MDN) #.43 
cor.test(all_years$PREDDEG,all_years$NPT4_PUB) #.38 
cor.test(all_years$PREDDEG,all_years$NPT4_PRIV) #.25 

#Learned that control has no effect on any of the variables I will be testing (always getting negative correlations)

#do after checking variable to make reading the df easier
head(all_years$CONTROL)
all_years$CONTROL <- factor(
  all_years$CONTROL,
  levels = c(1,2,3),
  labels = c("Public", "Private Nonprofit", "Private For-Profit")
)

all_years$PREDDEG <- factor(
  all_years$PREDDEG,
  levels = c(1,2,3,4),
  labels = c("Certificate", "Associate", "Bachelor's", "Graduate")
)

```

## Accounting for inflation then graphing ##
```{r}
#I forgot to include inflation with these graphs 
ggplot(all_years, aes(x = CONTROL, y = MD_EARN_WNE_P6)) + geom_boxplot()
ggplot(all_years, aes(x = CONTROL, y = MD_EARN_WNE_P10)) + geom_boxplot()
ggplot(all_years, aes(x = CONTROL, y = GT_25K_P10)) + geom_boxplot()

#formula to adjust for inflation ADD NPT4 for pub & priv into formula 
# inflation based on 2020/2021 
all_years <- all_years %>%
  mutate(
    inflation_adjust = case_when(
      year == "2009-2010" ~ 1.206,
      year == "2014-2015" ~ 1.093,
      year == "2020-2021" ~ 1.000
    ),
    NPT4_combined = coalesce(NPT4_PUB,NPT4_PRIV),
    
    MD_EARN_WNE_P6_real  = MD_EARN_WNE_P6 * inflation_adjust,
    MD_EARN_WNE_P10_real = MD_EARN_WNE_P10 * inflation_adjust,
    DEBT_MDN_real        = DEBT_MDN * inflation_adjust,
    NPT4_real = NPT4_combined * inflation_adjust,
    
    earn_to_debt_ratio = MD_EARN_WNE_P10_real / DEBT_MDN_real
  )

head(all_years)

# graph used to shown inflation over the years 
ggplot(all_years, aes(x = CONTROL, y = MD_EARN_WNE_P6_real, fill = year)) + 
  geom_boxplot(outlier.alpha = 0.2) + 
  labs( 
    title = "6 year post-entry median earnings by institution type",
    x = "Institution type",
    y = "Median earnings(inflation accounted to 2020 standards)"
  )

ggplot(all_years, aes(x = CONTROL, y = MD_EARN_WNE_P10_real, fill = year)) + 
  geom_boxplot(outlier.alpha = 0.2) + 
  labs( 
    title = "10 year post-entry median earnings by institution type",
    x = "Institution type",
    y = "Median earnings(inflation accounted to 2020 standards)"
  )

ggplot(all_years, aes(x = CONTROL, y = DEBT_MDN_real, fill = year)) + 
  geom_boxplot(outlier.alpha = 0.2) + 
  labs( 
    title = "median debt by institution type",
    x = "Institution type",
    y = "Median debt"
  )
# Plotting the threshold rate across years and school types #only shows 2009-2010 because other years don't provide this info
tail(all_years)

ggplot(all_years, aes(x = CONTROL, y = GT_25K_P10, fill = year)) +
  geom_boxplot(outlier.alpha = 0.2) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1), limits = c(0, 1)) +
  labs(
    title = "Share of Students Earning >$25k (10 Years Post-Entry)",
    x = "Institutional Control",
    y = "Percentage Earning >$25k",
  )

```
#grouping institute type & degree to compare the median data of selected variables 
```{r}
#remove PREDDEG to get a better understanding of numbers

#Median earning for 6 years
all_years%>%
   group_by(year,CONTROL) %>%
  summarize(med_earn = median(MD_EARN_WNE_P6_real, na.rm = TRUE))

all_years %>%
  group_by(year, CONTROL) %>%
  summarize(
    n_schools = sum(!is.na(MD_EARN_WNE_P6_real)),
    median_earnings_10yr = median(MD_EARN_WNE_P6_real, na.rm = TRUE),
  )

#Median earning for 10 years
all_years%>%
   group_by(year,CONTROL) %>%
  summarize(med_earn = median(MD_EARN_WNE_P10_real, na.rm = TRUE))

bachelor_schools %>% #now i can use this for a linear regression model
  group_by(year, CONTROL) %>%
  summarize(
    n_schools       = sum(!is.na(MD_EARN_WNE_P10_real)),
    med_earnings_10 = median(MD_EARN_WNE_P10_real, na.rm = TRUE),
    med_debt        = median(DEBT_MDN_real, na.rm = TRUE),
    med_net_price   = median(NPT4_real, na.rm = TRUE)
  )

all_years %>%
  group_by(year,CONTROL) %>%
  summarize(med_debt = median(DEBT_MDN_real, na.rm = TRUE))

all_years %>%
  group_by(year,CONTROL) %>%
  summarize(med_percent_past25k = median(GT_25K_P10, na.rm = TRUE))

bachelor_schools$DEBT_MDN_real

```
Regression models
```{r}

#creating a separate df that only hold bachelors to help answer my question
bachelor_schools <- all_years %>%
  filter(PREDDEG == "Bachelor's" & !is.na(MD_EARN_WNE_P10_real))

head(bachelor_schools)
gg_miss_var(bachelor_schools)
#looking at median income after 10 years & not great attempts
#model_md10<-lm(MD_EARN_WNE_P10_real ~ year, data = bachelor_schools)
model_md10_2<-lm(MD_EARN_WNE_P10_real ~ year + CONTROL, data = bachelor_schools)
model_realmd10<-lm(MD_EARN_WNE_P10_real ~ year * CONTROL, data = bachelor_schools)
model_ans_10<-lm(MD_EARN_WNE_P10_real ~ DEBT_MDN_real + year * CONTROL, data = bachelor_schools)

summary(model_md10_2)
summary(model_realmd10)
#Statistically significant but only explains around 12% of the variation so alot of unknown
summary(model_ans_10)

# good model model_md6<-lm(MD_EARN_WNE_P6_real ~ CONTROL, data = bachelor_schools) 
model_md6_2<-lm(MD_EARN_WNE_P6_real ~ year + CONTROL, data = bachelor_schools)
model_realmd6<-lm(MD_EARN_WNE_P6_real ~ year * CONTROL, data = bachelor_schools)
model_ans_6<-lm(MD_EARN_WNE_P6_real ~ DEBT_MDN_real + year * CONTROL, data = bachelor_schools)


summary(model_md6_2)
summary(model_realmd6)
#Statistically significant but only explains around 14% of the variation so alot of unknown
summary(model_ans_6)
```
#Answering/notes :how have post-graduation earnings changed over the same period?
-Overall Wages: After adjusting for inflation to constant 2020 dollars, post-graduation earnings did not steadily increase.
-2014/2015 dip: earnings dropped significantly during this time period due to many reason(look those up) but rebounded in 2020/2021(Covid year!)
-During 2020 the median earnings only increased by 312 for median earnings past 10 years but it's value isn't statistically significant at .72 but for 6 year it increased by 1,893$ & it's statistically significant
- Public & private nonprofit alum were able to maintain or increase their money over time while private for profit was losing money 

#"" block
