---
title: "EDA Project"
author: "Sydney"
date: "2026-09-20"
output: html_document
---
#Libraries
```{r}
#if you need to install a library use this line -- install.packages() 
library(readr)
library(tidyr)
```


##getting the data
```{r}
# put this code in your block n modify it to fit your needs like file path & columns

project.dir <- 
  "C:\Users\sydne\Desktop\DS_Fall26\DMungingExpAnalysis\Group Project 1"
# (change to your specific path up to 'Projects')
dataset.dir <- file.path(project.dir, "Datasets",
                         "College_Scorecard_Raw_Data_06102026", 
                         "College_Scorecard_Raw_Data_06032026") #the route to get to all the files
outputs.dir <- file.path(project.dir, "Output")
setwd(project.dir)
list.files(dataset.dir, pattern = "MERGED") #check to see if you're looking at the right files

year9_10 <- read_csv(
  file.path(dataset.dir, "MERGED2009_10_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)
year14_15 <- read_csv(
  file.path(dataset.dir, "MERGED2014_15_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)
year20_21 <- read_csv(
  file.path(dataset.dir, "MERGED2020_21_PP.csv"),
  col_select = c(""), #where we will select the columns we want to grab
)

```

#cleaning the data
```{r}
#we'll have to change all the PS values into 

```

#Reuel's block 
Columns important to my question "how have post-graduation earnings changed over the same period?"
# 1. Identifiers & Matching
  "UNITID",            # Unique institution ID
  "INSTNM",            # School name
  
  # 2. Institutional Type & Filtering (Apples-to-Apples)
  "CONTROL",           # 1 = Public, 2 = Private Nonprofit, 3 = Private For-Profit
  "PREDDEG",           # Predominant degree (3 = Predominantly Bachelor's)
  
  # 3. Post-Graduation Earnings (Primary Response Variables)
  "MD_EARN_WNE_P10",   # Median earnings of students working & not enrolled, 10 yrs post-entry
  "MD_EARN_WNE_P6",    # Median earnings of students working & not enrolled, 6 yrs post-entry
  "GT_25K_P10",        # Percentage earning > $25k (threshold of a high school graduate)
  
  # 4. Optional Context for Financial Investment (Cost & Debt)
  "DEBT_MDN",          # Median cumulative student debt at graduation
  "NPT4_PUB",          # Average net price for public institutions (Title IV students)
  "NPT4_PRIV"
```{r}
project.dir <- "C:/Projects"
# (change to your specific path up to 'Projects')
dataset.dir <- file.path(project.dir, "Datasets",
                         "College_Scorecard_Raw_Data_06102026", 
                         "College_Scorecard_Raw_Data_06032026") #the route to get to all the files
outputs.dir <- file.path(project.dir, "Output")
setwd(project.dir)
list.files(dataset.dir, pattern = "MERGED")

year9_10 <- read_csv(
  file.path(dataset.dir, "MERGED2009_10_PP.csv"),
   col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)
year14_15 <- read_csv(
  file.path(dataset.dir, "MERGED2014_15_PP.csv"),
  col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)
year20_21 <- read_csv(
  file.path(dataset.dir, "MERGED2020_21_PP.csv"),
 col_select = c("UNITID", "INSTNM", "CONTROL", "PREDDEG", "MD_EARN_WNE_P10","MD_EARN_WNE_P6","GT_25K_P10","DEBT_MDN","NPT4_PUB","NPT4_PRIV"),
  na = c("","NA","PrivacySuppressed")
)

head(year20_21)
```


Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
